Fixing Vulnerabilities Hinders Maintainability

### 2020-05-04 paper/icsme20.tex
review

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage{footnote}
\usepackage{xfrac}
\usepackage{balance}
\usepackage{hyperref}

@@

% \title{Fixing Vulnerabilities Hinders Maintainability}
\title{\Sof{TBA - New Titles in the comments}}
%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
%\title{The reason why you should be applying coding practices: safer software.}
%\title{What is the impact of security patches in software evolution?}
%\title{Hey, developers! You should pay more attention to your patches.}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1335$ security commits
\Sof{These are the initial values, should I use the final? In the end, 
we studied 1031 commits and 8 guidelines}. 
Results show 
that fixing software vulnerabilities may hinder the maintainability of 
open-source software. This is particularly the case for patches that 
address Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource 
Management Errors (CWE-399). We conclude that changes to code bases to patch 
vulnerabilities need to be performed with extra care and that security-related 
code evolution should be subject in computer curricula.
\end{abstract}

@@

\section{Introduction}
%
Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
Software quality characteristics include, but are not limited to functional correctness,
reliability, usability, maintainability, and security. Security is an 
important non-functional requirement during the development of software systems. 
However, there is still a considerable amount of vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate systems and 
exploit their resources. As an example, at the beginning of $2020$, academics published a 
new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
address in previous generations and that makes every user using a CPU released before Q$4$ $2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
at the cost of features and/or performance.
%

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010 considering
\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
However, many developers still lack the knowledge about best practices to deliver secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
supports the production of more reliable software preventing the introduction of 
faults/vulnerabilities in the code bases. ISO describes software 
maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
Despite these efforts, improving software security is not a trivial task and requires 
implementing patches that might affect software maintainability. 
Our suspiciousness is that some of the these patches may have a negative 
impact on the software maintainability and, possibly, even be the cause of the 
introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
in this paper, we present an empirical study on the impact of patches of 
vulnerabilities on software maintenance across open-source software.
%
% Software maintainability can also be defined as 
% the property of software that reveals how easily the software system can be 
% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
% Hedegus et al.
% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
% groups of refactored elements and non-refactored and conclude that the source
% code elements subjected to refactorings had significantly lower maintainability
% than the elements not modified.
% There is still little information about the impact of patching vulnerabilities
% in open-source software maintainability. \textit{Are developers affecting software
% maintainability when trying to improve software security?}, it is what we want to
% investigate.
%

As ISO does not provide any specific guidelines/formulas to calculate 
maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
code analysis service Better Code Hub (BCH) to compute the software compliance 
with a set of $10$ guidelines/metrics to produce quality software 
based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model). 
%
We use BCH's toolset to provide an assessment of maintainability in 
software mainly due to the broad number of metrics used.

From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability.\Sof{fix}Especially, for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Sof{-END-}This is important work
because there is still little information about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577}, documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
security-related code evolution in computer curricula.
%
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
%

The main contributions of the present work are:
%
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.
	\item A replication package with all scripts and data created to perform the
	empirical evaluation, for reproducibility. Available online:
  \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in the
protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and section~\ref{sec:discussion} discusses their
implications; section~\ref{sec:threats} enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software code base. However, these patches may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security patches impact
the maintainability of software.

%%%%% NEEDS REVIEW
As an example, consider the patch of the TLS state-machine protocol implementation in 
OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting TLS messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$~\footnote{CVE-$2016$-$6307$
details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
\emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
\url{https://github.com/openssl/openssl/commit/c1ef7c971d0bbf117c3c80f65b5875e2e7b024b1}
(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between the two parties. The problem in
CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
the memory for a message is allocated too early, prior to the excessive
message length check which can lead to a Denial of Service through memory
exhaustion if an attacker triggers up a certain message length to be allocated to service a
connection. In this case, there is only a security impact if the application does not call  \texttt{SSL\_free()} in a timely manner in the
event that the connection fails;
or, the application is working in a constrained environment where there
is very little free memory;
or, the attacker initiates multiple connection attempts such that there
are multiple connections in a state where memory has been allocated for
the connection, when \texttt{SSL\_free()} has not yet been called, and there is
insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
be transitory because as soon as the connection fails the memory is
subsequently freed again in the \texttt{SSL\_free()} call. However, there is an
increased risk during this period of application crashes due to the lack
of memory - which would then mean a more serious Denial of Service.

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ 
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
 int tls_get_message_header(SSL *s, int *mt){   
 // [snip]
  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
        SSL3_HM_HEADER_LENGTH;
- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-      goto err;
   }
 // [snip]

- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
-     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-     goto err;
- }

@@

The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
to handle the messages exchanged between peers. When messages are larger than
what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).

%
Although checking the maximum size of a message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the code base which produced a negative impact in the
maintainability. The developer tries to patch the
code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
and disrupts
one of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY} $-$
the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
OpenSSL maintainability.
\Sof{-- END --}
%%%%%%% 
Software maintainability is designed as the degree to which an application is understood, 
repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is 
approximately $75\%$ of the cost related to a project. To answer the following three 
research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
to measure the impact of
security patches on the maintainability of open-source software. 
%

\textit{\textbf{RQ1: What is the impact of security patches on the
maintainability of open-source software?}} Often security flaws require patching 
code to make software more secure.
However, \textbf{there is no evidence yet of how security patches impact the
maintainability of open-source software}. Our suspicion is that developers tend
to introduce technical debt in their software when patching the code base to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $1335$ commits using the
\emph{Better Code Hub} tool. 
%

\textit{\textbf{RQ2: Which patterns of security patches are more likely to
affect open-source software maintainability?}}

@@

the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to escape the data given by the variable

@@

\textit{\textbf{RQ3: What is the impact of security patches versus regular changes on the
maintainability of open-source software?}}

@@

We also compute the maintainability of random regular commits using the

@@

@@

@@

	Extract relevant data (e.g., owner and name of
	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and code bases from the combined dataset containing $1335$ security

@@

@@

the outcome of mining/manually inspect $265$ GitHub projects. Reis and Abreu 

@@

dataset of $711$ test cases for $16$ different security vulnerability types. Some of the vulnerability types

@@

code used to fix the security flaw. 

@@

for references to commits involved in the patch on the National Vulnerability Database (NVD) website. However, $70\%$

@@

used to patch the vulnerability. We assumed the last commit of the patch as 

@@

In this study, we focus on computing the

@@

$1031$ security patches. $912$ of these patches are identified with Common Weakness
Enumeration IDs (CWE) where $59.1\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $40.9\%$ were manually classified following the \texttt{Research Concepts} view map from the CWE's list.

% In Table~\ref{tab:patterns}, the most prevailing categories are presented. To satisfy the Wilcoxon test size requirement, types of vulnerabilities
% with less than $20$ instances were merged in a bigger group named \textit{Miscellaneous}.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage{footnote}
\usepackage{xfrac}
\usepackage{balance}
\usepackage{url}

@@

\title{Fixing Vulnerabilities Hinders Maintainability}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1037$ security commits. Results show 
that fixing software vulnerabilities hinders the maintainability of 
open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399). 
We conclude that changes to code bases to fix vulnerabilities need to be 
performed with extra care and that security-related code evolution should 
be subject in computer curricula.
\end{abstract}

@@

\section{Introduction}
Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
Software quality characteristics include, but are not limited to functional correctness,
reliability, usability, maintainability, and security. Security is an 
important non-functional requirement during the development of software systems. 
However, there is still a considerable amount of vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate systems and 
exploit their resources. \Sof{Improve example - outdated}\Luis{I think it is okay. Just say it is from 2019. It's too precise atm.}As an example, in the last week of January $2019$, 
a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
details available at \url{https://thehackernews.com/2019/01/apple-facetime-
privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
flaw that allowed callers to see and hear others without them picking up the call.%\cite{cacheOut}.
In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010 considering
\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
However, many developers still lack the knowledge about best practices to deliver secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
supports the production of more reliable software preventing the introduction of 
faults/vulnerabilities in the code bases. ISO describes software 
maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
Despite these efforts, improving software security is not a trivial task and requires 
implementing patches that might affect software maintainability. 
Our suspiciousness is that some of the these patches may have a negative 
impact on the software maintainability and, possibly, even be the cause of the 
introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
in this paper, we present an empirical study on the impact of patches of 
vulnerabilities on software maintenance across open-source software.
% Software maintainability can also be defined as 
% the property of software that reveals how easily the software system can be 
% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
% Hedegus et al.
% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
% groups of refactored elements and non-refactored and conclude that the source
% code elements subjected to refactorings had significantly lower maintainability
% than the elements not modified.
% There is still little information about the impact of patching vulnerabilities
% in open-source software maintainability. \textit{Are developers affecting software
% maintainability when trying to improve software security?}, it is what we want to
% investigate.
As ISO does not provide any specific guidelines or formulas to calculate 
maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
code analysis service Better Code Hub (BCH) to compute the software compliance 
with a set of $10$ guidelines/metrics to produce quality software 
based in ISO/IEC 25010~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model). 
We use BCH's toolset to provide an assessment of maintainability in 
software mainly due to the broad number of metrics used.
From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability. 
Especially, for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Luis{Little information is known about..}This is important work
because there is still little information about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577} and documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519}.
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
The main contributions of the present work are:
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.
	\item A replication package with all scripts and data created to perform the
	empirical evaluation, for reproducibility. Available online:
  \textit{<anonymized for double blind review>}.
This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in the
protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and section~\ref{sec:discussion} discusses their
implications; section~\ref{sec:threats}, enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
\section{Motivation and Research Questions}\label{sec:motivation}
Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software code base. However, these patches may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security patches impact
the maintainability of software.

%%%%% NEEDS REVIEW
As an example, consider the patch of the TLS state-machine protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting TLS messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$ \footnote{CVE-$2016$-$6307$
details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
\emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
\url{https://github.com/openssl/openssl/commit/c1ef7c971d0bbf117c3c80f65b5875e2e7b024b1}
(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between the two parties. The problem in
CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
memory for a message is allocated too early, prior to the excessive
message length check which can lead to a Denial of Service through memory
exhaustion if an attacker triggers up a certain message length to be allocated to service a
connection. In this case, there is only a security impact if the application does not call  \texttt{SSL\_free()} in a timely manner in the
event that the connection fails;
or, the application is working in a constrained environment where there
is very little free memory;
or, the attacker initiates multiple connection attempts such that there
are multiple connections in a state where memory has been allocated for
the connection, \texttt{SSL\_free()} has not yet been called, and there is
insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
be transitory because as soon as the connection fails the memory is
subsequently freed again in the \texttt{SSL\_free()} call. However there is an
increased risk during this period of application crashes due to the lack
of memory - which would then mean a more serious Denial of Service.

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
 int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
 // [snip]
  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
        SSL3_HM_HEADER_LENGTH;
- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-      goto err;
   }
 // [snip]

- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func5}!*/  
-     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-     goto err;
- }

@@

~\ref{lst:func1} \texttt{read\_state\_machine} function abstracts a sub-state-machine
to handle the messages exchanged between peers. When messages are larger than
what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
of this resource was being verified very early in the application's flow.  Listing ~\ref{lst:vuln2} presents those verifications (lines $5$-$7$ and line $11$-$13$).
These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}.

%
Although checking the maximum size of a message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the code base which produced a negative impact in the
maintainability. The developer tries to patch the
code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
and disrupts
one of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY} $-$
the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
OpenSSL maintainability.
%%%%%%% 
Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is approximately $75\%$ of the cost related to a project. To answer the following three research questions, we use two datasets
of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
security patches on the maintainability of open-source software. 
%

Often security flaws require patching code to make software more secure.
However, \textbf{there is no evidence yet of how security patches impact the
maintainability of open-source software}. Our suspicion is that developers tend
to introduce technical debt in their software when patching the code base to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $1037$ commits using the
\emph{Better Code Hub} tool. 
%
\subsubsection*{\textit{\textbf{RQ2}  \textbf{Which patterns of security patches are more likely to
affect open-source software maintainability?}}}

@@

the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to \textit{escape} the data given by the variable

@@

\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
maintainability of open-source software?}}}

@@

We also compute the maintainability of $1037$ random regular commits using the

@@

\Luis{this passive voice here sounds a bit weird (imo)}

@@

\Luis{could you make the numbers in the figure a bit larger}

@@

	Extract relevant data (e.g., sha, sha of the parent commit, owner of the project, name of 
	the repository) and code bases from the combined dataset containing $1335$ security

@@

\Luis{We already made clear that we ``evaluate the impact..'' }

@@

the outcome of mining/manually inspect $267$ GitHub projects\Luis{:mining and manually inspecting:}. \Luis{it is not clear the connection between the 267 projects and the works we mention below. There should be a sentence here clarifying and making the connection.} Reis and Abreu 

@@

dataset of $711$ test cases for $16$ security patterns. Some of the patterns

@@

code used to fix the security flaw. In this study, we focus on computing the

@@

for references to commits involved in the patch on the NVD's website. However, $70\%$

@@

used to patch the vulnerability. We assumed that the last commit of the patch as 

@@

@@

$1037$ security patches. $556$ of these patches are identified with Common Weakness
Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. Patterns
with less than $20$ instances were merged in a bigger pattern named \textit{Miscellaneous}.

\begin{table}[h]
\footnotesize
	\caption{Guidelines to produce maintainable code}
\begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}

\toprule
\textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
\midrule

\textbf{CWE-89} & 
 Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) & 
 When developers do not keep untrusted data
     separate from SQL queries. If an attacker sends a command that
     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule

\textbf{CWE-611} & 
 Improper Restriction of XML External Entity Reference (24 cases) & 
	XML documents are exchanged through the web containing entities with URIs
 	that resolve to local/external files. Thus, when XML parsers are not well configured,
 	attackers have allowed to directly those files.\\\midrule

 \textbf{CWE-20} & 
 Improper Input Validation (76 cases) & 
	When developers
 	forget to validate the inputs of an application, attackers may have control
 	of the control/data flow of the program.\\\midrule

  \textbf{CWE-200} & 
Information Exposure (39 cases)& 
Application's logs are one way of intentional/unintentional disclosure information
 	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule

   \textbf{CWE-352} & 
Cross-Site Request Forgery/CSRF (22 cases)& 
Poor session tokens generation
     and management usually allow attackers to send forged HTTP requests
     including authentication information from the victim to the vulnerable web
     application.\\\midrule

       \textbf{CWE-264} & 
Permissions, Privileges, and Access Controls (59 cases)& 
Incorrectly
 	    implemented functionalities related to authentication and session
 	    management, allowing an attacker to gain access to session tokens,
 	    passwords, keys, and other sensitive data.\\\midrule

 	           \textbf{CWE-399} & 
Resource Management Errors (22 cases)& 
Resource management issues are found more frequently
 		in programming languages that do not manage memory automatically (e.g., C/C++
 	    and Objective-C), i.e., where developers are responsible for
 	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule

 	     	           \textbf{CWE-79} & 
Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)& 
Lack of proper validation or escaping
 	    allows attackers to submit untrusted data to web browsers through malicious
 	    scripts that can hijack the user sessions or redirect the user to malicious
 	    sites.\\\midrule

 	     	           \textbf{CWE-22} & 
Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) & 
When developers forget to neutralize special elements within the path names to files/directories, attackers
 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
 	or entire repositories.\\\midrule

 	 	     	           \textbf{MISC} & 
Miscellaneous (673 cases) & 
This pattern comprises
 		several other security refactorings that do not have a CWE assigned and patterns that
 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
\bottomrule
\end{tabular}
\label{tab:patterns}
\end{table}
